---
title: 'PS239T: California’s Investment in PK-12 Public School Infrastructure'
author: "Lindsay Maple"
date: "Fall 2016"
output: ioslides_presentation
---
```{r, include=FALSE, error=TRUE}
library(choroplethr)
library(choroplethrMaps)
library(leaflet)
library(sp)
library(RColorBrewer)
library(foreign)
library(dplyr)
library(maps)
library(mapdata)
library(rgdal)
library(maptools)
library(rgeos)
library(readstata13)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r data cleaning, echo=TRUE, error=TRUE, include=FALSE}
setwd(dir="~/Documents/CCS_Housing/")

district_data <- read.csv("masterdata_LM.csv")

#Need stata package to read
alldata_MO <- read.dta13("alldata12.dta")

alldata_subset <- subset(alldata_MO,
                         select=c(cds_dist,dist_name,county,bc,bcstudent,sfp_nc,sfp_mod,sfp_fh,
                         mo5yr,cap5yr,opsp5yr,sfpyears_mo_ps,sfpyears_cap_ps,sfpyears_opsp_ps,
                         sfpyears_statecap_ps,currdebt,currdebt_ps,sf_total,rep_value,localcap,localcap5yr,
                         gf_benchmark,crv_4pct,crv_3pct,crv_2pct,benchmark_gap,mo_gfbenchmark_gap,mo_gap,
                         cap_gap))


district_subset <- subset(district_data,
                          select=c(cds_id,fips_code,District..Name,County.Name,AV.Per.Student,FRPM.Share,
                                   Total.Enrollment))

##Change the names to merge the datasets
names(district_subset) <- c("cds_dist","fips_code","dist_name", "county", "avperstudent", "frpm_share", "total_enroll") 

merged_data <- merge(alldata_subset, district_subset, by=c("cds_dist", "county", "dist_name"), all.x=TRUE)
```

## California School Facilities Spending

The majority of school districts in California do not meet minimum annual facility expenditure benchmarks, even—in many cases—with state funding. 

Between 2008 and 2012, at least 57% of districts did not meet industry benchmarks for spending on capital renewals and more than 60% failed to meet the benchmark for basic maintenance and operations. 

In many cases, the same school districts are falling behind on both measures.


## Project Design

- This project required lots of data cleaning and merging. I used Stata for some of the data cleaning before merging datasets in R. 
- Choroplethr provided the basis for understanding where and how much spending is going towards public school facilities. 
- Leaflet enabled me to include pop-ups for interactive maps that show specifics about each California county.


```{r AVmap, include=FALSE, error=TRUE}
#subsetting data so I'm not dealing with a million variables
county.spending <- subset(district_data, 
                          select=c(fips_code, County.Name, Bond.Passed.11.2016,Bond.Passed.6.2016, 
                                   AV.Per.Student, FRPM.Share, Region.Code))

#fips codes for Choropleth requirements
fipscodes1 <- data.frame(region=county.spending$fips_code, value=county.spending$AV.Per.Student)

avfipsclean <- fipscodes1[complete.cases(fipscodes1),]

fips.av <- avfipsclean %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 

##Choropleth map for Assessed value
AVmap <- county_choropleth(fips.av, 
                  title = "Assessed Value Per Student by County",
                  legend = "AV $ per student", 
                  num_colors = 5,
                  state_zoom = "california")

#data frames by names for Leaflet requirements
countynames <- data.frame(region=county.spending$County.Name, value=county.spending$AV.Per.Student)

avnamesclean <- countynames[complete.cases(countynames),]

names.av <- avnamesclean %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 
```

```{r, error=TRUE, include=FALSE}
## Choropleth: Average Assessed Value per Student by County 
AVmap
```


```{r FRPM Map, echo=TRUE, error=TRUE, include=FALSE}
#fips codes for Choropleth requirements
FRPM_df <- data.frame(region=district_data$fips_code, value=district_data$FRPM.Share)

clean_FRPM<-FRPM_df[complete.cases(FRPM_df),]

CA.FRPM <- clean_FRPM %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 

## Choropleth map for Free & Reduced Lunch
FRPMmap <- county_choropleth(CA.FRPM, 
                  title = "Free & Reduced Priced Lunch by County",
                  legend = "FRPM Average \nProportion by County", 
                  num_colors = 6,
                  state_zoom = "california")

#data frames by county names for Leaflet requirements
FRPMnames <- data.frame(region=merged_data$county, value=merged_data$frpm_share)

frpmnamesclean <- FRPMnames[complete.cases(FRPMnames),]

names_frpm <- frpmnamesclean %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 
```

```{r, error=TRUE, include=FALSE}
## Choropleth: Average Free & Reduced Priced Lunch Share by County 
FRPMmap
```

```{r M&O spending, echo=TRUE, error=TRUE, include=FALSE}
#fips codes for Choropleth requirements
MOspending_df <- data.frame(region=merged_data$fips_code, value=merged_data$mo_gap)

clean_MOspdf<-MOspending_df[complete.cases(MOspending_df),]

MOspending <- clean_MOspdf %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 

#Choropleth map for gap in spending above/below benchmark
MOmap <- county_choropleth(MOspending, 
                  title = "Gap in M&O Benchmark Spending \nby County",
                  legend = "Dollars Away from Benchmark",
                  num_colors = 6,
                  state_zoom = "california") 

# data frames by county names for Leaflet requirements
MOnames <- data.frame(region=merged_data$county, value=merged_data$mo_gap)

clean_MOnames <- MOnames[complete.cases(MOnames),]

MOnames_final <- clean_MOnames %>% 
  group_by(region)  %>% 
  summarise(value=mean(value, na.rm=T)) 
```


```{r, error=TRUE, include=FALSE}
## Choropleth: Maintenance & Operations Spending by County 
MOmap
```


```{r AV interactive map, error=TRUE, include=FALSE}
#input shapefile for California counties 
counties <- readShapeSpatial('CA_counties.shp')
names(counties)
sort(counties$NAME)

#merging in the data into this shapefile
#datacounties <- data.frame(counties@data, names.av[match(counties[,"NAME"], names.av[,"region"]),])

#set color palette
pal <- colorNumeric("YlGn", NULL, n = 5)

#create the popup text
county_popup <- paste0("<strong>County: </strong>", 
                      names.av$region, 
                      "<br><strong>AV per student:</strong>", 
                      names.av$value)

#write the interactive Leaflet map
assessedvalue <- leaflet(counties) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(fillColor = ~pal(names.av$value),
              weight = 1,
              stroke = TRUE,
              fillOpacity = .8,
              data = counties,
              popup = ~paste("<strong>County:</strong>",NAME,
                             "<br>",
                             "<strong>AV per student:</strong>",names.av$value,
                             "<br>")) %>%
  addLegend(title = "Assessed Value <br> Per Student", 
            values = names.av$value, 
            bins=3, 
            labels = 5,
            position="topright", 
            pal = pal,
            labFormat = labelFormat(suffix = ''))
```

## Leaflet: Assessed Value Per Student Interactive Map
```{r AV, error=TRUE}
assessedvalue
```

```{r FRPM interactive map, error=TRUE, include=FALSE}
#create the popup text
FRPM_popup <- paste0("<strong>County: </strong>", 
                      names_frpm$region, 
                      "<br><strong>Average FRPM Share:</strong>", 
                      names_frpm$value)

#set color palette
FRPMpal <- colorQuantile("YlGn", NULL, n = 5)

#write the interactive Leaflet map
FRPMleaflet <- leaflet(counties) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(fillColor = ~FRPMpal(names_frpm$value),
              weight = 1,
              stroke = TRUE,
              fillOpacity = .8,
              data = counties,
              popup = ~paste("<strong>County:</strong>",NAME,
                             "<br>",
                             "<strong>Average FRPM Share:</strong>",names_frpm$value,
                             "<br>")) %>%
  addLegend(title = "Average FRPM Share", 
            values = names_frpm$value, 
            bins=3, 
            labels = 5,
            position="topright", 
            pal = FRPMpal,
            labFormat = labelFormat(suffix = ''))
```

## Leaflet: FRPM Interactive Map by County
```{r, error=TRUE}
FRPMleaflet
```

```{r MO interactive map, error=TRUE, include=FALSE}
#Set color palette for this map
MOpal <- colorNumeric(
  palette = "Greens",
  domain = MOnames_final$value)

#write the interactive Leaflet map
MOleaflet <- leaflet(counties) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(fillColor = ~MOpal(MOnames_final$value),
              weight = 1,
              stroke = TRUE,
              fillOpacity = .8,
              data = counties,
              popup = ~paste("<strong>County:</strong>",NAME,
                             "<br>",
                             "<strong>Current spending <br> from benchmark:</strong>",MOnames_final$value,
                             "<br>")) %>%
  addLegend(title = "Current Spending from <br>Minimum Benchmark", 
            values = MOnames_final$value, 
            bins=10, 
            labels = 5,
            position="topright", 
            pal = MOpal,
            labFormat = labelFormat(suffix = ''))
```

## Leaflet: Maintenance & Operations Spending Gap by County
```{r, error=TRUE}
MOleaflet
```

## Next Steps

- There are still some flaws in the code (ie. on the Assessed Value leaflet map, I have no idea where those values are coming from; and similarly for the Benchmark Spending)
- I want to add more layers to the leaflet maps including which counties passed bond measures in 2016 for school infrastructure. 
- I couldn't get the Choropleth maps to knit into this HTML, so I decided to remove them. However, I think they can be rather helpful (even though I prefer the Leaflet maps)
